A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-07-08 faculty.tex
services

\chapter{Preface}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students} (and also domestic ones), especially those in  smaller countries or less well-known universities, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and universities in other countries).

In addition, this document can help \textbf{US faculty and admission committee} gain a better understanding of international students and their cultural differences.  By recognizing and leveraging these differences, CS programs in the US can attract larger and more competitive application pools from international students.

I wish you the best of luck. And if you follow this guidance, you will at least have a good chance at GMU (see
\href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU}{why
  you want to study at here}). Happy school hunting!

@@

\begin{center}
  \href{https://nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf}{nguyenthanhvuh.github.io/phd-cs-us/demystify.pdf},
\end{center}

@@

\subsection{Advice} To serve, you can directly ask people to recommend you. For example, for PC, ask senior colleagues or even your PhD advisor/collaborators to recommend you to the PC chairs. You can also directly email the PC chairs and ask to serve.  Once you have served well, you will likely be invited to serve again.  
Similarly, for NSF Panels, you can cold email program managers in your area and ask to be on their panels.  Unfortunately, I do not have much experience with non-NSF panels or proposals. For leadership roles, you can directly ask the general chair of a conference to see if you can take more leadership roles.

\subsection{Personal Experience} In my case I mostly cold email people for opportunities. For example, I directly email OOPSLA and PLDI PC chairs and asked, and they accepted my request. I tried this with SE conferences  e.g., ICSE (and also nominate myself through the official form), and it never works, even though I have many papers at ICSE. I also got invited to serve on a PC because a senior colleague recommended me.
For NSF, I emailed program managers and asked to serve on their panels.

I felt that I didn't get much out of serving on PCs and didn't really like the review process much. However, I still do it because I feel it is my responsibility, i.e., people spent time reviewing my papers, so I should also volunteer my time to review theirs.

On the other hand, I thoroughly enjoy serving on NSF panels (though of course my proposals still got rejected frequently). While the work is more intense and takes more time, I found it more rewarding and enjoy the discussions. Since my first NSF panel in 2019, I have served 8 panels (at least once every year since '19) and I look forward to serving more.

I have also volunteered for leadership roles, e.g., I directly asked the general chair of ICSE'21 to see if I can take more leadership roles and was then invited to serve as a proceeding co-chair. In hindsight this particular role is a lot of work, and I didn't really like many aspects of it, but it was an eye-openning experience.

\section{Service to the Department}
Your department chair will assign you to some committees to help run the department.  For tenure-track faculty the services are often lighter than tenured ones (e.g., less leadership roles). Common committees include PhD and/or MS admission, faculty search, and colloquium committees.

@@

\subsection{Advice} You should try to be strategic about which committees you serve on that can help build your lab and tenure case. For example, serving in the PhD adcom is very common for new assistant professors, as it is a good way to recruit students for your lab. Serving on the faculty search committee is also a good way to recruit new faculty members that you like (and also let you know what the job market is like, what is hot and not). Serving on the colloquium committee is also a good way to invite speakers in your field that can be your \emph{tenure letter writers}.

@@

\section{Service to the University}
In most cases, these are optional services and assistant professors often do not take any of them.  Common examples include serving the faculty senate, scholarship committee, etc. The university often sends out emails asking for volunteers (so somewhat generic and you can safely ignore), and sometimes people (e.g., your dept chair or dean) will directly ask you to serve (in which case you will want to consider it).

\subsection{Advice and Personal Experience} I only have limited experience with these services. I served in a couple of grants and scholarship committees, in which I review applications for scholarship and grants for students. I also help represent the college in serving as judge of the KimmyDuong foundation scholarship. These are in general time-consuming (e.g., the KimmyDuong scholarship reviews and interview process happen during weekends!), but they are also rewarding as you get to know students and help them. In general, I would recommend doing these services only if you have the time.  However, sometimes they can be rewarding and fun, so you might want to try them out.

\section{Other Types of Services}
\paragraph{Writing Recommendation Letters} You will be asked to write recommendation letters for both students and colleagues.

\textbf{Advice}: %you should have a template for recommendation letters.  You can find many templates online.  You should also ask your colleagues to share their templates with you.  You should also ask your colleagues to share their recommendation letters with you.  You can learn a lot from reading other people's recommendation letters.

>>>>

\chapter{Preface}
This documents my tenure track journey in computer science and the lessons I have learned along the way. It is a way for me to reflect on my experiences and share them with others who may be interested in pursuing a similar path. I hope that it will be helpful to those who are in their early stages of academia (e.g., tenure-track faculty), especially in computer science.

@@

\begin{center}
  \href{https://nguyenthanhvuh.github.io/phd-cs-us/faculty.pdf}{nguyenthanhvuh.github.io/phd-cs-us/faculty.pdf},
\end{center}

@@

\subsection{How to get invited for these services?}

You can directly ask people to recommend you. For example, for PC, ask senior colleagues or even your PhD advisor/collaborators to recommend you to the PC chairs. You can also directly email the PC chairs and ask to serve.  Once you have served well, you will likely be invited to serve again.  
Similarly, for NSF Panels, you can cold email program managers in your area and ask to be on their panels.  Unfortunately, I do not have much experience with non-NSF panels or proposals. For leadership roles, you can directly ask the general chair of a conference to see if you can take more leadership roles.

\subsection{Personal Experience} I mostly cold email people for opportunities. For example, I directly email OOPSLA and PLDI PC chairs and asked, and they invited. I tried this with ICSE (and also nominate myself through the official form), and it never works, even though I have many papers at ICSE (that leaves a bad taste in my mouth about ICSE, but that is another story).  I also got invited to serve on a PC because a senior colleague recommended me.
For NSF, I just emailed program managers and asked to serve on their panels, and they invited me.

I felt that I didn't get much out of serving on conf. PCs and didn't really like the review process much. 
However, I still do it because I feel it is my responsibility, i.e., people spent time reviewing my papers, so I should also volunteer my time to review theirs.

On the other hand, I thoroughly enjoy serving on NSF panels (though of course my proposals still got rejected very often). To this date I have served on 8 NSF panels (at least once every year since 2019) and I look forward to serving more. While NSF review is more intense and takes more time (compared to conf. reviews), I found it more rewarding and enjoy the discussions.

I have also volunteered for leadership roles, e.g., I directly asked the general chair of ICSE'21 to see if I can take more leadership roles and was then invited to serve as a proceeding co-chair. In hindsight this particular role is a lot of work, and I didn't really like some aspects of it, but it was an eye-openning experience.

\section{Service to the Department}
Your department chair will assign you to some committees to help run the department.  For tenure-track faculty the services are often lighter than tenured ones (e.g., less leadership roles).

Common committees include PhD and/or MS admission, faculty search, and colloquium committees. Serving in the PhD adcom is very common for new assistant professors, as it is a good way to recruit students for your lab. Serving on the faculty search committee is also a good way to recruit new faculty members that you like (and also let you know what the job market is like, what is hot and not). Finally, being on the colloquium committee is a good way to invite speakers in your field that can be your \emph{tenure letter writers}.

@@

@@

\section{Service to the University}
You can also serve the university in various committees.
Common examples include serving the faculty senate, scholarship committee, etc. The university often sends out emails asking for volunteers (so somewhat generic and you can safely ignore), and sometimes people (e.g., your dept chair or dean) will directly ask you to serve (in which case you will want to consider it).

\subsection{Personal Experience} I have limited experience with these services. I served in a couple of grants and scholarship committees, in which I review applications for scholarship and grants for students. I also help represent the college in serving as judge of the Kimmy-Duong foundation scholarship. These are time-consuming (e.g., the KimmyDuong scholarship reviews and interview process happen during weekends!), but they are also rewarding as you get to know students and help them (many have unique and interesting stories).

In general, I would recommend doing these services only if you have the time and have interest in them. However, sometimes they can be rewarding and fun, so you might want to try them out.

\section{Other Types of Services}

\paragraph{Writing Recommendation Letters} You will be asked to write recommendation letters for both students and sometimes even colleagues. For example, I write on average ten reference letters for students each year, including letters for grad schools, jobs, or scholarships. I view writing letters as my responsibility as a faculty member, and so I usually do not say no to any student who asks for a letter. However, I would tell a student I do not know well that I can only write a generic  (i.e., weak) letter for them.

\section{Conclusion}

Many would advise about being strategic about your services.I think it is a good advice as you can get more out of your services. However, I do not follow this, and in many cases I would volunteer for tasks that no one wants to do because I feel that someone gotta do it and I can help out.
